---
title: "Retrieving JASPAR databases through AnnotationHub"
output: BiocStyle::html_document
author:
    name: Damir Baranasic
    affiliation: Division of electronics,
bibliography: JASPAR.bib
vignette: >
  %\VignetteIndexEntry{Retrieving JASPAR databases through AnnotationHub}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

JASPAR (https://jaspar.elixir.no/) is an open-access database that has provided
high-quality, manually curated, and non-redundant DNA binding profiles for
transcription factors (TFs) as position frequency matrices (PFMs) for over 20
years. For this 11th release of the database, we expanded
our CORE and UNVALIDATED collections with new PFMs and updated existing
profiles. Specifically, we added 306 and 433 new binding profiles to the CORE (a
12% increase) and UNVALIDATED (a 60% increase) collections, respectively, and
updated 13 profiles from the CORE collection. With the new collections of PFMs,
we updated the TF binding site predictions in eight species and provided the
corresponding genomic tracks. Moreover, we updated the profile clusters and
familial TF binding sites accordingly. This release introduces a regulatory
sequence simulation tool, inMOTIFin, enabling users to simulate regulatory
sequences using JASPAR PFMs. To enrich the annotations linked to the TFs,
JASPAR now provides scientific literature-based human TF target information
derived from a large language model. Notably, this release features the
addition of a new collection of deep learning BPnet models, providing a
paradigm shift in modeling and characterizing TF-DNA interactions, driven by
advances in artificial intelligence. JASPAR 2026 provides 1304 BPNet models
trained on Homo sapiens ENCODE ChIP-seq datasets from 264 TFs. Interpretation of
the BPnet revealed critical binding profiles that were curated and matched to
JASPAR CORE PFMs for an easy overview of the binding patterns. Profiles from
BPnet models associated with the same TF were clustered to provide a
non-redundant summary of the binding properties of the TF, resulting in 266
primary motifs in the new Deep Learning collection. The 2026 release of JASPAR
now offers the classical manually curated collections of TF binding profiles and
state-of-the-art deep learning models. These collections lay a foundation for
future endeavors in genomic research, serving the scientific community in
uncovering the mechanisms of gene regulation.

The *JASPAR* package gives access to the JASPAR databases from R. It does not
contain any data. The databases are SQLite files that are registered as
resources in `r Biocpkg("AnnotationHub")`. The package finds the resource of the
release you ask for, lets AnnotationHub download it once, and returns the path
of the cached file. You can open this file with `r CRANpkg("RSQLite")` or use it
with `r Biocpkg("TFBSTools")` [@Tan:2016]. The default release is JASPAR 2024
[@Rauluseviciute:2024]. JASPAR 2022 [@10.1093/nar/gkab1113] can be retrieved as
well.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("JASPAR")
```

# JASPAR in AnnotationHub

Start by connecting to AnnotationHub and searching it for JASPAR.

```{r hub, message = FALSE}
library(AnnotationHub)
library(JASPAR)

ah <- AnnotationHub()
query(ah, "JASPAR")
```

The search also finds resources that only mention JASPAR in their title, such as
genomic tracks. The JASPAR databases are the resources whose R data class is
`JASPAR`. Each of them is one release, and the title is the version name. The
accession is the identifier of the resource in AnnotationHub.

```{r records}
records <- ah[ah$rdataclass == "JASPAR"]
data.frame(accession = names(records),
           version = records$title,
           added = records$rdatadateadded)
```

`getAvailableJASPARVersions()` returns the same version names, newest first. The
package takes them from the AnnotationHub metadata. It does not contain a list
of its own, so a release that is added to AnnotationHub, or whose record is
corrected, can be retrieved without updating the package.

```{r versions}
getAvailableJASPARVersions(hub = ah)
```

A listed version can still fail to download when the location registered for it
in AnnotationHub cannot be reached. The error then names the version and the
accession, see [Troubleshooting](#troubleshooting).

# Retrieving a database

`JASPAR()` retrieves a database by its version. The default version is
`JASPAR2024`. The first call of `AnnotationHub()` downloads the hub database
(about 140 MB). Every later call takes a few seconds, so create the hub once, as
above, and pass it with `hub = ah` when you retrieve several versions.

```{r retrieve}
jaspar <- JASPAR(version = "JASPAR2024", hub = ah)
version(jaspar)
```

`db()` returns the path of the SQLite file in the AnnotationHub cache. The file
is downloaded the first time and reused afterwards.

```{r path}
file.exists(db(jaspar))
```

This is the same file that AnnotationHub returns for the accession of the
resource.

```{r accession, message = FALSE}
id <- names(records)[records$title == "JASPAR2024"]
identical(unname(ah[[id]]), db(jaspar))
```

# Searching profiles with TFBSTools

`getMatrixSet()` of `r Biocpkg("TFBSTools")` reads profiles from a JASPAR SQLite
file. Open the file read-only, so that the cached copy cannot be changed by
accident, and give the connection to `getMatrixSet()`. The options select the
profiles. This query returns the CORE profiles of the budding yeast
*Saccharomyces cerevisiae* (NCBI taxonomy identifier 4932).

```{r tfbstools, message = FALSE}
library(RSQLite)
library(TFBSTools)

con <- dbConnect(SQLite(), db(jaspar), flags = SQLITE_RO)
pfms <- getMatrixSet(con, list(species = 4932, collection = "CORE"))
dbDisconnect(con)

length(pfms)
head(sapply(pfms, name))
```

The result is a `PFMatrixList`. Each element is a position frequency matrix with
the identifier and the name of the profile.

```{r matrix}
pfm <- pfms[[1]]
c(ID = ID(pfm), name = name(pfm))
Matrix(pfm)
```

`getMatrixSet()` also accepts the path of the file. It then opens and closes the
connection itself, for reading and writing, so the read-only connection above is
the safer way to open the cached file.

```{r path-method}
pfms <- getMatrixSet(db(jaspar), list(species = 4932, collection = "CORE"))
length(pfms)
```

# Comparing releases

Each release is a separate database, so you can compare them. The following code
counts the profiles of JASPAR2022 and JASPAR2024 in each collection with SQL. A
profile can have several versions, which are separate rows of the `MATRIX`
table, so the code counts each base identifier once.

```{r compare}
count_profiles <- function(release) {
    jaspar <- JASPAR(version = release, hub = ah)
    con <- dbConnect(SQLite(), db(jaspar), flags = SQLITE_RO)
    on.exit(dbDisconnect(con))
    counts <- dbGetQuery(con,
        "SELECT COLLECTION AS collection,
                COUNT(DISTINCT BASE_ID) AS profiles
         FROM MATRIX GROUP BY COLLECTION")
    cbind(release = release, counts)
}

do.call(rbind, lapply(c("JASPAR2022", "JASPAR2024"), count_profiles))
```

# Working offline

AnnotationHub needs an internet connection, also to create the hub object.
Downloaded files stay in the cache. To use them without a connection, create the
hub with `localHub = TRUE`. This hub contains only the resources that were
downloaded before, so only those versions can be retrieved.

```{r offline, eval = FALSE}
ah_local <- AnnotationHub(localHub = TRUE)
JASPAR(version = "JASPAR2024", hub = ah_local)
```

# Troubleshooting {#troubleshooting}

If a database cannot be retrieved, the error names the version and the
accession, and gives the reason reported by the server (for example
`HTTP 404 Not Found`). A failed download is not stored in the cache, so there is
nothing to clean up: try again later. You do not have to clear the cache either
after the location of a resource has been corrected in AnnotationHub, the next
call retrieves the file.

AnnotationHub does not look into the files it caches. The package therefore
checks that the file starts like an SQLite database and is not shorter than its
header says, and downloads it once more if it is not, for example after an
interrupted transfer. A hub created with
`localHub = TRUE` cannot download, so with such a hub the error asks you to
retrieve the version with an online hub. You can also ask for a fresh copy
yourself.

```{r force, eval = FALSE}
AnnotationHub::cache(ah[id], force = TRUE)
```

# Session Info

Here is the output of `sessionInfo()` on the system on which this document was
compiled:

```{r session_info, echo = FALSE}
sessionInfo()
```

# References
